Regional Effects of Marine Protected Area Networks Are Unclear

### 2020-04-18 documents/ovando-regional-effects-of-mpas-v2.Rmd
fixed some package version issues

---
title: Regional Effects of Marine Protected Area Networks Are Unclear
author: 
  - Daniel Ovando
  - Jennifer E. Caselle
  - Christopher Costello
  - Olivier Deschenes
  - Steven D. Gaines
  - Ray Hilborn
  - Owen Liu
abstract: |
   Marine Protected Areas (MPAs) cover between 3-7% of the world's oceans, up from less than 1\% in the year 2000. The Convention on Biological Diversity calls for 10% of coastal waters to be protected inside MPAs by 2020, while the International Union for Conservation of Nature calls for 30\% protection by 2030. It is often clear that MPAs produce conservation benefits inside their borders, but many MPAs are also justified on the grounds that they also benefit the broader region outside their borders. The conservation effects of MPAs are most commonly evaluated using response ratios---measures of biomass densities inside their borders, relative to biomass densities in reference sites outside their borders. Studies of this nature have provided broad evidence that MPAs produce conservation benefits within their borders. While these gains can be vitally important, marine populations are rarely contained entirely within the borders of MPAs; therefore a critical question is not only do MPAs produce conservation gains inside their borders, but how do they affect the broader region in which they are located?  The Channel Islands National Marine Sanctuary provides a clear example of this challenge. A network of MPAs covering roughly 20\% of the Islands’ waters was put in place in 2003, with a goal of providing regional conservation and fishery benefits. Response ratios from the region indicate that the Channel Island MPAs have increased biomass densities inside the MPAs. However, we are unable to find a clear effect of these same MPAs at the regional scale across multiple species.Building off of existing theory, we use a bio-economic simulation model to explain this discrepancy, and demonstrate under what conditions we likely can and cannot clearly estimate the regional conservation effects of MPAs. We show that MPA networks covering 25\% or less of a region are likely to produce regional increases in fish biomass densities on the order of 25\% or less, a meaningful effect but also one that can easily be overwhelmed by environmental shocks. Our results provide a novel assessment of the regional effects of a large and iconic Marine Protected Area network, and provide guidance for managers charged with monitoring and adapting MPAs. 
bibliography: My Library.bib
csl: pnas.csl
output:
  bookdown::word_document2: default
  bookdown::pdf_document2: default
  bookdown::html_document2: default
params:
  run_name: ["v6.0"]
  sim_years: [50]
  burn_years: [20]
  num_patches: [25]
linkcolor: blue
toc: false
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}\linenumbers
  - \usepackage{draftwatermark}
---

@@

## significance
  Healthy marine ecosystems are critical to the well-being of the planet. Marine protected areas, parts of the oceans protected from human activities such as fishing, are increasingly being used in an effort to conserve and manage these ecosystems. Our study pairs theory with empirical methods to examine the regional effects of MPAs on conservation outcomes. Using simulations, we show that the regional effects of marine protected areas can be highly variable and dependent on human drivers. We find no statistically clear effect of a network of MPAs after 13 years of protection, which our simulation analysis suggests is actually to be expected. Our results lay a foundation for future research on the design and monitoring of marine protected areas.

## acknowledgements
  Funding for the simulation model was provided by the NMFS-Sea Grant Population and Ecosystem Dynamics Fellowship. This study would not be possible without the work provided by PISCO divers over the years. DO thanks Cody Szuwalski, Julia Lawson, and André Punt for helpful comments and technical support.

@@

As stakeholders around the world increasingly seek to use MPAs in marine resource management portfolios, it is critical that we develop a better understanding of the magnitude and drivers of regional-scale MPA effects. This is particularly true in the context of a changing climate, in which "set it and forget it" management strategies are increasingly untenable; we need to be able to monitor the performance of MPAs in order to adapt them as dictated by a shifting environment. The MPAs within the Northern Channel Islands National Marine Sanctuary, California, USA (which we will refer to as the Channel Islands from now on) provides an ideal case study to address this need. A network of protected areas covering approximately 20% of the Islands' waters was put in place in 2003 [see @osmond2010, @kirlin2013, @botsford2014, and @hilborn2012 for information on the creation of these MPAs]. This MPA network has subsequently been used as a model case study in protected area design around the world XX.

We use data from the first 14 years of protection to provide what is to our knowledge the first empirical assessment of the regional effect of a large MPA network on a wide array of finfish species. In contrast to clear differences in biomass densities observed inside and outside of well protected MPAs both globally [@lester2009] and in the Channel Islands [@caselle2015] we are unable to detect a clear regional effect from the Channel Islands MPAs. We build off of existing MPA theory to consider why this might be, and provide guidance for scientists and managers as to when and how we might expect to estimate the regional conservation effects of MPAs.

## Results

@@

We define the regional conservation effects of MPAs as the change in total biomass densities of fish relative to the total biomass densities of fish that would have occurred without the MPAs (acknowledging that other outcomes such as increased biodiversity or resiliency are also important to conservation but are beyond the scope of this analysis). Numerous factors can affect the regional effects of MPAs. These include the scale of adult and larval dispersal relative to the size of the MPAs  [@gaines2003; @botsford2008; @difranco2018;@mcgilliard2008;@gerber2005], larval dispersal patterns and the strength and timing of density dependence in the population (e.g. pre- or post-settlement, @burgess2014), how overfished the population would be without the MPA, and how fishing and management responds to the implementation of the MPAs [@hilborn2004; @hilborn1992; @walters2004; @hastings2003;@gerber2003; @hilborn2004a; @gaines2010; @botsford2003;@smith2003a;@smith2006;@walters2000a;@white2011]. In addition, even for the same total area of MPAs, the location and spacing of the MPAs can have a profound influence on their cumulative impact through habitat and network effects [@gaines2010; @costello2010a].

@@

## Discussion {#discussion .unnumbered}

@@

While we do not have access to fine scale fishing data from the Northern Channel Islands alone, reported catches for the species of interest in the Santa Barbara region in fact exhibit an overall downward trend in the years post reserve (see Fig.SXX). We can most likely rule a negative MPA effect caused by a constant-catch fishing strategy then.  What is another possible explanation for the recent downward trend in the estimated MPA effects? The Channel Islands region (and the entire West coast of the USA) experienced a dramatic ‘marine heatwave’ beginning in 2014 and persisting through 2016, resulting in part in extremely elevated water temperatures throughout the region  [@gentemann2017]. Biogeographic differences in the distributions of targeted and non-targeted species may confound the observed effects of MPAs. Many of the non-targeted species have warm thermal affinities and have increased in numbers since the heatwave [@freedman2019]. The targeted group is made up mostly of fishes with cold-water affinities, such as members of the genus *Sebastes*. As such, we hypothesize that the recent evidence for a decline in densities of targeted species is due to environmental conditions that disproportionately affect the targeted group (and not for example due to concentrated fishing pressure outside the reserves). This hypothesis is supported by the sharp declines in densities of targeted species seen inside the MPAs themselves. If the cause of recent declines was due to increases in fishing pressure we would expect to see substantial declines only in the fished areas (Fig.\@ref(fig:raw-density)).

@@

Containing a carefully designed, well-enforced, and well-studied MPA network, The Channel Islands would seem at face value to be an ideal location to study the regional conservation effects of protected areas. The persistently high response ratios suggest that despite overall decreases in targeted biomass densities inside and outside MPAs, the MPAs may still be providing protection within their borders. But, as we have shown here these response ratios are not necessarily an indicator of regional effects. The difference-in-difference strategy utilized here presents an alternative identification strategy, that while not without its own strict caveats presents some potential improvements over response-ratios as a means of estimating regional conservation effects. While we estimate a highly uncertain but overall positive effect at first, we are unable to detect a robust signal from 2012-2017. After 14 years of MPA protection we are left without a clear picture of the effect of the Channel Island MPA network on biomass densities of targeted fin-fish species.

Our simulation results suggest though that we should not be surprised by this result. The Channel Islands MPAs cover approximately 20% of the waters in the Channel Islands, and while formal stock assessments are not available for many of the targeted species in our analysis, what evidence we have suggests that, as a group, these fish are on average not heavily overfished. Some species, such as California sheephead and blue rockfish were likely below target levels during the period [@alonzo2004;@dick2017], but projections based upon the overall average response across all species will likely suggest modest benefits even if a subset of species could experience much larger population gains. Harkening back to our simulations, we expect the average percentage difference in densities of targeted species with and without MPAs to be modest (Fig.\ref{fig:mpa-effect}). Effects of this size are likely to be challenging to detect empirically given the large natural variation of marine ecosystems (especially temperate reefs) and the observation error inherent in infrequent annual survey programs such as those provided by PISCO.

@@

As an additional complication, our simulations and difference-in-difference model examine percentage changes in biomass densities. As an extreme example, an increase in biomass densities from .2kg/m^2^ to .4kg/m^2^ would be produce as a 100% increase. While a large percentage increase, this is a small change in absolute biomass densities relative to the variance in the observation process itself (see Fig.S1 for a companion to Fig.\ref{fig:sim-effects} scaled by absolute population size). In addition, the median estimated age at sexual maturity for the targeted species included in this study is `r median_tm` years, meaning that the span of this analysis represents less than three generations of MPA protection for half of the the measured species. Ongoing monitoring may yet reveal clearer effects. Analysis of more rapidly growing and maturing species, e.g. spiny lobster, may also reveal clearer signals.

Given the natural variability of marine ecosystems, and the strong challenges of obtaining accurate samples from oceanic environments, how large of an effect would an MPA network have to allow a difference-in-difference strategy such as this to be a reliable measure of MPA effects? To provide guidance on this important question, we used our bio-economic model to simulate data from a range of scenarios with increasing MPA effect size, along with increasing degrees of observation error and natural recruitment variation. As an added measure, we include scenarios in which the sampled species go through recruitment regimes, which may be positive for both targeted and non-targeted species, or positive for non-targeted species and negative for targeted species. We then used a simple Bayesian difference-in-difference regression on these simulated data, and estimated the percent error between the posterior probability distribution of the estimated MPA effect and true simulated MPA effect.

While unbiased across simulations, the difference-in-difference model struggled severely when MPA effect sizes were less than 25% and the model was faced with observation and process errors (Fig.\@ref(fig:val-plot)). Even larger effect sizes commonly misestimated the true MPA effect by 50% or more.  Obtaining a mean absolute percent error (MAPE) of 25% or less across our simulated datasets required an MPA effect of at least 30%. This is merely an illustrative exercise, omitting clearly critical factors such as detection probability and sampling strategy. However, since nearly any omission which one can think of would make an MPA effect harder to detect, not easier, these results serve as a useful floor for the likely difficulty in estimating MPA effects. In the context of the Channel Islands, given the potential effect size produced by out simulation model these results suggest that we might expect to be unable to detect a clear effect.

@@

```{r val-plot, include = TRUE,fig.width = width_1col, fig.asp=asp_1col, fig.cap = "Distribution of percent error in posterior estimate of MPA effect (y-axis) plotted against true simulated MPA effect (x-axis). Color shows concentration of simulations. Black line shows mean absolute percent error as a function of true simulated MPA effect."}

@@

Suppose that we are willing to tolerate a MAPE of 25%. Our analysis suggests that we would need an MPA effect size of at least 30% to achieve this. Two of the most critical drivers of MPA performance are the size of the MPA and the degree of fishing pressure. Looking across these two variables, our simulations suggest if the MPA network covers 25% or more of a species range and/or pre-MPA depletion is greater than 60% we might expect an effect size with our target MAPE. While recently some extremely large MPAs have been enacted that may indeed reach into the higher levels of MPA coverage, for near-shore commercial fin-fish many MPA networks are likely to cover areas more in line with the Channel Islands (20%), and as such have effect sizes that may be difficult to detect (Fig.\@ref(fig:sim-effects)-A).

These initial simulation results would seem to suggest some rather simple rules of thumb: put an MPA of sufficient size on an at least somewhat overexploited population and one can expect large and potentially detectable results. However, the MPA literature highlights a large number of variables beyond simply size and fishing pressure that can affect performance. To address this, we examined the variability in expected MPA effects across each of the two major axes (pre-MPA depletion and MPA size). As both MPA size and pre-MPA depletion increase, the potential for a large MPA effect increases. However, for both variables even at extremely large or extremely small values a wide range of MPA effects were possible (though as we might expect the effect of pre-MPA depletion, in other words fishing pressure, was much clearer than MPA size alone) (Fig.\@ref(fig:sim-effects)-B).

To put these results into context, the FAO estimates that 7% of the worlds fisheries with status estimates fall into the relatively unexploited category (roughly depletion less than 50%), 60% fall into the fully fished category (roughly depletion 50%-70%), and 33% fall into the heavily fished category (roughly depletion greater than 70%) [@fao2018, assuming a Bmsy/B0 ratio of 40%, per @thorson2012], though works that include a broader range of fisheries estimate that 50% or more of stocks to fall into the heavily fished category [@costello2012;@rosenberg2018]. The regional effects of MPA networks covering 25% or less of a species range may be difficult to detect in many places with already well-managed fisheries, while for that size we might expect clearer effects in less-managed locations (though that of course ignores the complication of compliance with MPA regulations). Within these broad guidelines though a wide range of outcomes are possible: as a starting place users can use the bio-economic MPA simulation model developed for this paper to explore potential outcomes for specific MPAs using an interactive web application available at [danovando.shinyapps.io/simmpa](https://danovando.shinyapps.io/simmpa/).

@@

MPAs are an important part of the marine resource management toolbox. Under ideal circumstances they can protect individual species and ecosystem linkages, while supporting local economies through tourism and fishing opportunities. One rationale for the expansion of MPAs is that they will deliver net conservation benefits both inside *and outside* their borders. To date, this assumption is insufficiently tested, and this is the focus of our paper. Our results show that regional conservation benefits of MPAs are highly context dependent and in many circumstances, are likely to be small enough that they are nearly impossible to detect empirically. Indeed, this is exactly what we found in our empirical case study from the large and well-studied network of MPAs in the Channel Islands, California, USA.

@@

Once simulation models have been used to help design an MPA, how can users evaluate whether it is achieving their objectives? Response ratios are commonly used as evidence for conservation outcomes of MPAs; @lester2009 and @halpern2002 present meta-analyses of hundreds of such studies. These results often find massively higher densities and biomass inside MPAs than outside [median biomass increases of  over 400%, @lester2009]. But as suggested in @white2011, @ferraro2018, @larsen2019 and further demonstrated here, without careful attention to the design of control sites (accounting for example for the displacement of fishing effort by MPAs), response ratios may be highly unreliable estimators of regional MPA effects. When MPAs affect nearby control sites used in response ratios through biological spillover or concentration of fishing effort, it is entirely possible to for MPA to produce massive response ratios while simultaneously having minimal effects on the entire population partially protected by the MPAs. As @larsen2019 suggests, there are many potential alternatives for estimating the effects of MPAs that better account for the challenges of causal inference (though that may be more data-intensive). We applied one such approach here (a difference-in-difference estimator), and yet were still unable to reach robust conclusions as to the effect of MPAs on the biomass density of targeted finfish in the Channel Islands, due to the likely small size of the true effect relative to the strength and variability of environmental drivers.

While this does not mean that all MPAs will face similar challenges in estimating their effects, our results in the rigorously studied and iconic Channel Islands Marine Protected Area network make clear that in many instances empirically detecting a clear regional effect of MPAs may not be possible. How then should stakeholders go about adaptively monitoring and managing MPAs? Simulation modeling can help inform the range of effect sizes that may be expected, and monitoring programs can perhaps be tuned to focus on the species groups that have the highest chance of a detectable effect size over the early years of the reserve [@nickols2019]. Expanding data collection to include robust monitoring of spatio-temporal fleet dynamics may help assess the validity of control sites used in response ratios, support the direct inclusion of these fleet dynamics into statistical models, and allow managers to take into account potential negative interactions between MPAs and fleet dynamics such as those that may occur under constant-catch dynamics. Whenever possible monitoring programs should be implemented prior to MPA implementation to provide a pre-treatment benchmark. Non-equilibrium analyses also help set expectations for effect sizes over time [@nickols2019,Fig.\ref{did-plot}]. Educating communities about the challenges of estimating the effects of MPAs can help set expectations, so that a lack of a clear effect is not necessarily viewed as a failure of the program, or large positive result based solely on response ratios as a clear sign of success and subsequent increased fishing opportunity. Rather, results and subsequent management actions (such as adaptation of MPA network design) must be considered in the context of reasonable expectations for the MPAs in question. While this paper has focused on the conservation outcomes of MPAs, future work must also address the challenge of predicting and estimating the fishery impacts of protected areas.

As the number and size of global MPA networks increase, it is critical that we both set appropriate expectations for their outcomes, and plan how we will monitor the performance of these protected areas over time. While the history of MPA science has made important strides in helping us understand the dynamics of protected areas, the future of MPA science must directly tackle the challenge of evaluating the performance of these MPAs at the regional scale, and adapting their design as needed to best achieve objectives. This is particularly true if communities are depending on MPAs as their primary marine resource management tool. Commonly employed metrics such as response ratios may be applicable in some circumstances, but can have severe shortcomings as metrics of regional conservation effects. Dependence on unreliable estimators of MPA effects may lead to stakeholders incorrectly attributing negative environmental shocks as MPA failures, or interpreting data arising from scorched-earth fishing outside MPAs as a conservation success. Bio-economic modeling can help frame community expectations, reducing the potential for a reduction in support if unrealistic conservation or fishery expectations are not realized. Statistical approaches that explicitly address complications such as the spatial spillover effects of MPAs (such as the difference-in-difference approach used here) may give users an improved understanding of the performance of their MPAs, but even they may struggle when expected effect sizes are small. Clearly communicating what we should expect, and what we can detect, from MPAs is critical in ensuring that MPAs play effective roles in fisheries management and marine conservation.

>>>>

---
title: Regional Effects of Marine Protected Area Networks Are Unclear
author: 
  - Daniel Ovando
  - Jennifer E. Caselle
  - Christopher Costello
  - Olivier Deschenes
  - Steven D. Gaines
  - Ray Hilborn
  - Owen Liu
abstract: |
   Marine Protected Areas (MPAs) cover 3-7% of the world's oceans, up from less than 1\% in the year 2000. The Convention on Biological Diversity called for 10% of coastal waters to be protected inside MPAs by 2020, while the International Union for Conservation of Nature is now calling for 30\% protection by 2030. It is often clear that MPAs produce conservation benefits inside their borders, but many MPAs are also justified on the grounds that they may benefit the broader region outside their borders. The conservation effects of MPAs are most commonly evaluated using response ratios---measures of numerical or biomass densities inside their borders, relative to these metrics in reference sites outside their borders. Studies of this nature have provided broad evidence that MPAs produce conservation benefits within their borders. While these gains can be vitally important, marine populations are rarely contained entirely within the borders of MPAs; therefore a critical question is not only do MPAs produce conservation gains inside their borders, but how do they affect the broader region in which they are located?  The Channel Islands National Marine Sanctuary provides a clear example of this challenge. A network of MPAs covering 20\% of the Islands’ waters was put in place in 2003, with a goal of providing regional conservation and fishery benefits. Response ratios from the region indicate that the Channel Island MPAs have increased biomass densities inside the MPAs. We are unable to find a clear effect of these same MPAs on biomass densities throughout the Channel Islands region. We use a bio-economic simulation model to explain this discrepancy, and demonstrate under what conditions we likely can and cannot clearly estimate the regional conservation effects of MPAs. MPA networks covering 25\% or less of a region are likely to produce regional increases in fish biomass densities on the order of 25\% or less, a meaningful effect but also one that can easily be overwhelmed by environmental shocks. Our results provide a novel assessment of the regional effects of a large and iconic Marine Protected Area network, and provide guidance for managers charged with monitoring and adapting MPAs. 
bibliography: My Library.bib
csl: pnas.csl
output:
  bookdown::word_document2: default
  bookdown::pdf_document2: default
  bookdown::html_document2: default
params:
  run_name: ["v6.0"]
  sim_years: [50]
  burn_years: [20]
  num_patches: [25]
linkcolor: blue
toc: false
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}\linenumbers
  - \usepackage{draftwatermark}
---

@@

## significance
  Healthy marine ecosystems are critical to the well-being of the planet. Marine protected areas, parts of the oceans protected from human activities such as fishing, are increasingly being used to conserve and manage these ecosystems. Our study pairs theory with empirical methods to examine the regional effects of MPAs on conservation outcomes. We show that the regional effects of marine protected areas can be highly variable and dependent on human drivers. We find no statistically clear effect of a network of MPAs after 14 years of protection, which our simulation analysis suggests is actually to be expected. Our results lay a foundation for future research on the design and monitoring of marine protected areas.

## acknowledgements
  Funding for the simulation model was provided by the NMFS-Sea Grant Population and Ecosystem Dynamics Fellowship. Empirical data collection was funded primarily by the David and Lucille Packard Foundation in support of the Partnership for Interdisciplinary Studies of Coastal Oceans (PISCO) with additional funding from the California Ocean Protection Council and California SeaGrant. This study would not be possible without the work provided by PISCO divers over the years. DO thanks Cody Szuwalski, Julia Lawson, and André Punt for helpful comments and technical support and special thanks to Kathyrn Davis Koehn and Avrey Parsons-Field.

@@

As stakeholders around the world increasingly seek to use MPAs in marine resource management portfolios, it is critical that we develop a better understanding of the magnitude and drivers of regional-scale MPA effects. This is particularly true in the context of a changing climate, in which "set it and forget it" management strategies are increasingly untenable; we need to be able to monitor the performance of MPAs in order to adapt them as dictated by a shifting environment. The MPAs within the Channel Islands National Marine Sanctuary, California, USA (which we will refer to as the Channel Islands from now on) provides an ideal case study to address this need. A network of protected areas covering approximately 20% of the Islands' waters was put in place in 2003 [see @osmond2010, @kirlin2013, @botsford2014, and @hilborn2012 for information on the creation of these MPAs]. This MPA network has subsequently been used as a model case study in protected area design around the world [@botsford2014 and references therein].

We use data from the first 14 years of protection to provide what is to our knowledge the first empirical assessment of the regional effect (defined here as the effect of MPAs on the mean biomass density of targeted finfish both inside and outside of MPAs) of a large MPA network on a wide array of finfish species. In contrast to clear differences in biomass densities observed inside and outside of well protected MPAs both globally [@lester2009] and in the Channel Islands [@caselle2015] we are unable to detect a clear regional effect from the Channel Islands MPAs. We build off of existing MPA theory to consider why this might be, and provide guidance for scientists and managers as to when and how we might expect to estimate the regional conservation effects of MPAs.

## Results & Discussion

@@

We define the regional conservation effects of MPAs as the change in mean total biomass densities of targeted finfish both inside and outside of MPAs, relative to the mean total biomass densities of targeted finfish inside and outside of MPAs that would have occurred without the MPAs (acknowledging that other outcomes such as increased biodiversity or resiliency are also important to conservation but are beyond the scope of this analysis). Consider a evenly distributed population that has 50% of its range protected by an MPA. Suppose that the MPA increase biomass densities inside the reserves by 20%, and by 0% outside the reserve. The regional effect of the MPA would be 10%.

Numerous factors can affect the regional effects of MPAs. These include the scale of adult and larval dispersal relative to the size of the MPAs  [@gaines2003; @botsford2008; @difranco2018;@mcgilliard2008;@gerber2005], larval dispersal patterns and the strength and timing of density dependence in the population (e.g. pre- or post-settlement, @burgess2014), how overfished the population would be without the MPA, and how fishing and management responds to the implementation of the MPAs [@hilborn2004; @hilborn1992; @walters2004; @hastings2003;@gerber2003; @hilborn2004a; @gaines2010; @botsford2003;@smith2003a;@smith2006;@walters2000a;@white2011]. In addition, even for the same total area of MPAs, the location and spacing of the MPAs can have a profound influence on their cumulative impact through habitat and network effects [@gaines2010; @costello2010a].

@@

@@

While we do not have access to fine scale fishing data from the Northern Channel Islands alone, reported catches for the species of interest in the Santa Barbara region in fact exhibit an overall downward trend in the years post reserve (see Fig.SXX). We can most likely rule out a negative MPA effect caused by a constant-catch fishing strategy then.  What is another possible explanation for the recent downward trend in the estimated MPA effects? The Channel Islands region (and the entire West coast of the USA) experienced a dramatic ‘marine heatwave’ beginning in 2014 and persisting through 2016, resulting in part in extremely elevated water temperatures throughout the region  [@gentemann2017]. Biogeographic differences in the distributions of targeted and non-targeted species may confound the observed effects of MPAs. Many of the non-targeted species in the Channel Islands have warm thermal affinities and have increased in numbers since the heatwave [@freedman2019]. The targeted group is made up mostly of fishes with cold-water affinities, such as members of the genus *Sebastes*. As such, we hypothesize that the recent evidence for a decline in densities of targeted species is due to environmental conditions that disproportionately affect the targeted group (and not for example due to concentrated fishing pressure outside the reserves). This hypothesis is supported by the declining trend in mean biomass densities of targeted species seen inside the MPAs themselves. If the cause of recent declines was due to increases in fishing pressure we would expect to see substantial declines only in the fished areas (Fig.\@ref(fig:raw-density)).

@@

Containing a carefully designed, well-enforced, and well-studied MPA network, The Channel Islands would seem at face value to be an ideal location to study the regional conservation effects of protected areas. The persistently high response ratios suggest that despite overall decreases in targeted biomass densities inside and outside MPAs, the MPAs may still be providing protection within their borders. But, as we have shown here these response ratios are not necessarily an indicator of regional effects. The difference-in-difference strategy utilized here presents an alternative identification strategy, that while not without its own strict caveats presents some potential improvements over response-ratios as a means of estimating regional conservation effects. While we estimate a highly uncertain but overall positive effect at first, we are unable to detect a robust signal from 2012-2017. We believe that the dissapearance of increasing probabilities of positive MPA effects estimated by our model are likely driven by the marine heatwave experienced by the region beginning in 2014. After 14 years of MPA protection and following a large environmental perturbation we are left without a clear picture of the effect of the Channel Island MPA network on biomass densities of targeted fin-fish species.

Our simulation results suggest that we should not be surprised by this result. The Channel Islands MPAs cover 20% of the surface waters in the Channel Islands, and while formal stock assessments are not available for many of the targeted species in our analysis, what evidence we have suggests that, as a group, these fish are on average not heavily overfished. Some species, such as California sheephead and blue rockfish were likely below target levels during the period [@alonzo2004;@dick2017], but projections based upon the overall average response across all species will likely suggest modest benefits even if a subset of species could experience much larger population gains. Our simulations suggest that the average percentage difference in densities of targeted species with and without MPAs to be modest (Fig.\ref{fig:mpa-effect}). Effects of this size are likely to be challenging to detect empirically given the large natural variation of marine ecosystems (especially temperate reefs) and the observation error inherent in infrequent (e.g. annual) common in marine systems.

@@

As an additional complication, our simulations and difference-in-difference model examine percentage changes in biomass densities (as oppose to changes in total population size). As an extreme example, an increase in biomass densities from 0.2kg/m^2^ to 0.4kg/m^2^ would be produce as a 100% increase. While a large percentage increase, this is a small change in biomass densities relative to the variance in the observation process itself (see Fig.S1 for a companion to Fig.\ref{fig:sim-effects} scaled by absolute population size). In addition, the median estimated age at sexual maturity for the targeted species included in this study is `r median_tm` years, meaning that the span of this analysis represents less than three generations of MPA protection for half of the the measured species. Ongoing monitoring may yet reveal clearer effects. Analysis of more rapidly growing and maturing species, e.g. spiny lobster, may also reveal clearer signals.

Given the natural variability of marine ecosystems, and the large challenges of obtaining accurate samples from oceanic environments, how large of an effect would an MPA network have to have in order to allow a difference-in-difference strategy such as this to be a reliable measure of MPA effects? To provide guidance on this important question, we used our bio-economic model to simulate data from a range of scenarios with increasing MPA effect size, along with increasing degrees of observation error and natural recruitment variation. As an added measure, we include scenarios in which the sampled species go through recruitment regimes, which may be positive for both targeted and non-targeted species, or positive for non-targeted species and negative for targeted species. We then used a simple Bayesian difference-in-difference regression styled after the full model used here on these simulated data, and estimated the percent error between the posterior probability distribution of the estimated MPA effect and true simulated MPA effect.

While unbiased across simulations, the difference-in-difference model struggled severely when MPA effect sizes were less than 25% and the model was faced with observation and process errors (Fig.\@ref(fig:val-plot)). Even larger effect sizes commonly misestimated the true MPA effect by 50% or more.  Obtaining a mean absolute percent error (MAPE) of 25% or less across our simulated datasets required an MPA effect of at least 30%. This is merely an illustrative exercise, omitting clearly critical factors such as detection probability and sampling strategy. However, since nearly any omission which one can think of would make an MPA effect harder to detect, not easier, these results serve as a useful floor for the likely difficulty in estimating MPA effects. In the context of the Channel Islands, given the potential effect size produced by our simulation model these results suggest that we might expect to be unable to detect a clear effect.

@@

```{r val-plot, include = TRUE,fig.width = width_1col, fig.asp=asp_1col, fig.cap = "Distribution of percent error in posterior estimate of MPA effect (y-axis) plotted against true simulated MPA effect (x-axis). Color shows concentration of simulations. Black line shows mean absolute percent error (MAPE) as a function of true simulated MPA effect."}

@@

Suppose that we are willing to tolerate a MAPE of 25%. Our analysis suggests that we would need an MPA effect size of at least 30% to achieve this. Two of the most critical drivers of MPA performance are the size of the MPA and the degree of fishing pressure. Looking across these two variables, our simulations suggest if the MPA network covers 25% or more of a species range and pre-MPA depletion is greater than 60% we might expect an effect size with our target MAPE. While recently some extremely large MPAs have been enacted that may indeed reach into the higher levels of MPA coverage, for near-shore commercial fin-fish many MPA networks are likely to cover areas more in line with the Channel Islands (20%), and as such have effect sizes that may be difficult to detect (Fig.\@ref(fig:sim-effects)-A).

These initial simulation results suggest some rather simple rules of thumb: put an MPA of sufficient size on an at least somewhat overexploited population and one can expect large and potentially detectable results. The MPA literature highlights a large number of variables beyond simply size and fishing pressure that can affect performance. To address this, we examined the variability in expected MPA effects across each of the two major axes (pre-MPA depletion and MPA size). As both MPA size and pre-MPA depletion increase, the potential for a large MPA effect increases. However, for both variables even at extremely large or extremely small values a wide range of MPA effects were possible (though as we might expect the effect of pre-MPA depletion, in other words fishing pressure, was much clearer than MPA size alone) (Fig.\@ref(fig:sim-effects)-B).

To put these results into context, the FAO estimates that 7% of the worlds fisheries with status estimates fall into the relatively unexploited category (roughly depletion less than 50%), 60% fall into the fully fished category (roughly depletion 50%-70%), and 33% fall into the heavily fished category (roughly depletion greater than 70%) [@fao2018, assuming a biomass at maximum sustainable yield to unfished biomass ratio of 40%, per @thorson2012], though works that include a broader range of fisheries estimate that 50% or more of stocks to fall into the heavily fished category [@costello2012;@rosenberg2018]. The regional effects of MPA networks covering 25% or less of a species range may be difficult to detect in many places with already well-managed fisheries, while for that size we might expect clearer effects in less-managed locations (though that of course ignores the complication of compliance with MPA regulations). Within these broad guidelines a wide range of outcomes are possible based on local fleet and fish dynamics: as a starting place users can use the bio-economic MPA simulation model developed for this paper to explore potential outcomes for specific MPAs using an interactive web application available at [danovando.shinyapps.io/simmpa](https://danovando.shinyapps.io/simmpa/).

@@

MPAs are an important part of the marine resource management toolbox. Under ideal circumstances they can protect individual species and ecosystem linkages, while supporting local economies through tourism and fishing opportunities. One rationale for the expansion of MPAs is that they will deliver net conservation benefits both inside *and outside* their borders. To date, this assumption is insufficiently tested, and this is the focus of our paper. Our results show that regional conservation benefits of MPAs are highly context dependent and in many circumstances, are likely to be small enough that they are nearly impossible to detect empirically. Indeed, this is exactly what we found in our empirical case study from the relatively large and well-studied network of MPAs in the Channel Islands, California, USA.

@@

Once simulation models have been used to help design an MPA, how can users evaluate whether it is achieving their objectives? Response ratios are commonly used as evidence for conservation outcomes of MPAs; @lester2009 and @halpern2002 present meta-analyses of hundreds of such studies. These results often find massively higher densities and biomass inside MPAs than outside [median biomass increases of  over 400%, @lester2009]. But as suggested in @white2011, @ferraro2018, @larsen2019 and further demonstrated here, without careful attention to the design of control sites (e.g. accounting for the displacement of fishing effort by MPAs), response ratios may be highly unreliable estimators of regional MPA effects. When MPAs affect nearby control sites used in response ratios through biological spillover or concentration of fishing effort, it is entirely possible to for MPA to produce massive response ratios while simultaneously having minimal effects on the entire population partially protected by the MPAs. As @larsen2019 suggests, there are many potential alternatives for estimating the effects of MPAs that better account for the challenges of causal inference (though that may be more data-intensive). We applied one such approach here (a difference-in-difference estimator), and yet were still unable to reach robust conclusions as to the effect of MPAs on the biomass density of targeted finfish in the Channel Islands, due to the likely small size of the true effect relative to the strength and variability of environmental drivers.

While this does not mean that all MPAs will face similar challenges in estimating their effects, our results in the rigorously studied and iconic Channel Islands Marine Protected Area network make clear that in many instances empirically detecting a clear regional effect of MPAs may not be possible. How then should stakeholders go about adaptively monitoring and managing MPAs? Simulation modeling can help inform the range of effect sizes that may be expected, and monitoring programs can perhaps be tuned to focus on the species groups that have the highest chance of a detectable effect size over the early years of the reserve [@nickols2019]. Expanding data collection to include robust monitoring of spatio-temporal fleet dynamics may help assess the validity of control sites used in response ratios, support the direct inclusion of these fleet dynamics into statistical models, and allow managers to take into account potential negative interactions between MPAs and fleet dynamics such as those that may occur under constant-catch dynamics. Whenever possible monitoring programs should be implemented prior to MPA implementation to provide a pre-treatment benchmark.

Non-equilibrium analyses also help set expectations for effect sizes over time [@nickols2019,Fig.\ref{did-plot}]. Educating communities about the challenges of estimating the effects of MPAs can help set expectations, so that a lack of a clear effect is not necessarily viewed as a failure of the program, or large positive result based solely on response ratios as a clear sign of success and subsequent increased fishing opportunity. Rather, results and subsequent management actions (such as adaptation of MPA networks) must be considered in the context of reasonable expectations given the size, age, and degree of enforcement of the MPAs in question, together with the ecological and economic dynamics of a given system. While this paper has focused on the conservation outcomes of MPAs, future work must also address the challenge of predicting and estimating the fishery impacts of protected areas.

As the number and size of global MPA networks increase, it is critical that we both set appropriate expectations for their outcomes, and plan how we will monitor the performance of these protected areas over time. While the history of MPA science has made important strides in helping us understand the dynamics of protected areas, the future of MPA science must directly tackle the challenge of evaluating the performance of these MPAs at the regional scale, and adapting their design as needed to best achieve objectives. This is particularly true if communities are depending on MPAs as their primary marine resource management tool. Commonly employed metrics such as response ratios may be applicable in some circumstances, but can have severe shortcomings as metrics of regional conservation effects. Dependence on unreliable estimators of MPA effects may lead to stakeholders incorrectly attributing negative environmental shocks as MPA failures, or interpreting data arising from scorched-earth fishing outside MPAs as a conservation success. Both of these scenarios would hinder the ability of MPAs to serve as effective marine resource management tools. Bio-economic modeling can help frame community expectations, reducing the potential for a reduction in support if unrealistic conservation or fishery expectations are not realized. Statistical approaches that explicitly address complications such as the spatial spillover effects of MPAs (such as the difference-in-difference approach used here) may give users an improved understanding of the performance of their MPAs, but even they may struggle when expected effect sizes are small. Clearly communicating what we should expect, and what we can detect, from MPAs is critical in ensuring that MPAs play effective roles in fisheries management and marine conservation.
